E-Graphs With Bindings

### 2025-01-23 LICS_2025/e_graphs_with_bindings/sections/combinatorial_semantics.tex
last section left for closed

First,  we fix some notation.  Let $(-)^*$ be the free monoid monad over \textsf{Set}.  
We extend this to the case of relations $R \subseteq V \times V$,  so that we denote by $R^{*} \subseteq V^* \times V^*$ the element-wise extension of $R$ to a relation on ordered sequences. 
We will sometimes use functional notation for relations,  so that given $R \subseteq V \times W$ and $v \in V$,  $R(v) \subseteq W$. 
We elide associativity and unit isomorphisms associated with coproducts $(+)$,  and denote by $\iota_j: X_{j} \rightarrow X_{1} + \ldots X_{j} + \ldots X_{n}$ the $j^{\text{th}}$ injection,  and $[f,g]: A_1 + A_2 \to B$ the co-pairing of two morphisms $f: A_1 \to B$ and $g:A_2 \to B$. 
We denote by $A +_{f,g} B$ the pushout of the span $A \xleftarrow{f} C \xrightarrow{g} B$.
We will also refer to $A,B$ as \textit{feet} of the cospan, and to $C$ as \textit{carrier}.

@@

In ~\cite{bonchi_string_2022-1},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for categorical presentations of SMTs including a Frobenius algebra. 
Hypergraphs generalise graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams (\textit{i.e.}, without requiring a Frobenius algebra) and the category of (cospans of) such hypergraphs.

\begin{definition}[Category of hypergraphs]\label{def:hypergraph}
A \emph{hypergraph $\mathcal{G}$ over a signature $\Sigma$} is a tuple $(V,E,s,t,l)$,  where $V$ is a finite set of vertices, $E$ is a finite set of edges, $s : E \to V^{*}$ is a source function, $t : E \to V^{*}$ is a target function,  and $l : E \to \Sigma$ is a labelling function that assigns each edge a generator from monoidal signature $\Sigma$.  The labelling function must respect the typing of the generator: for all edges $e$,  $|s(e)| = m$ and $|t(e)| = n$,  where $l(e) : m \to n$.  We call a hypergraph \textit{discrete} if its set of edges is empty.   
A \emph{hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ is given by a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that the following hold. 
\begin{enumerate}
    \item $\phi_V^*(s_{\mathcal{F}}(e)) = s_{\mathcal{G}}(\phi_E(e))$
    \item $\phi_V^*(t_{\mathcal{F}}(e)) = t_{\mathcal{G}}(\phi_E(e))$
    \item $l_{\mathcal{F}}(e) = l_{\mathcal{G}}(\phi_E(e))$
\end{enumerate}
We denote by $\catname{Hyp(\Sigma)}$ the category of hypergraphs over $\Sigma$ and hypergraph homomorphisms. 
\end{definition}
Note that $\catname{Hyp(\Sigma)}$ has all finite colimits and, in particular,  the coproduct is given by the disjoint union of hypergraphs.
The initial object is given by the empty hypergraph.

\begin{definition}[Symmetric monoidal category of cospans]
Let $\mathbb{C}$ be a category with all finite colimits.  A cospan from $X$ to $Y$ is a pair of arrows $X \xrightarrow{} A \xleftarrow{} Y$  in $\mathbb{C}$.  This \textit{category of cospans of $\mathbb{C}$},  denoted $\catname{Csp(\mathbb{C})}$,  has as objects those of $\mathbb{C}$ and as arrows isomorphism classes of cospans.
%Two cospans $X \xrightarrow{} A \xleftarrow{} Y$ and $X \xrightarrow{} B \xleftarrow{} Y$ are isomorphic if the following diagram commutes and $\alpha$ is iso.
%\[
%\begin{tikzcd}
%                                               & A \arrow[dd, "\alpha"] &                                                \\
%X \arrow[ru, bend left] \arrow[rd, bend right] &                        & Y \arrow[ld, bend left] \arrow[lu, bend right] \\
%                                               & B                      &                                               
%\end{tikzcd}
%\]
%For $X \in \mathbb{C}$ an $id$-cospan is $X \xrightarrow{id_X} X \xleftarrow{id_X} X$.
Composition of cospans $X \xrightarrow{f} A \xleftarrow{g} Y$ and $Y \xrightarrow{h} B \xleftarrow{i} Z$ is given by pushout: $X \xrightarrow{f} A +_{g,h} B \xleftarrow{i} Z$,  with identity given by the cospan of identities.  Its monoidal product is given by the coproduct of $\mathbb{C}$ as follows: 
\ifdefined \ONECOLUMN
\[
    (A \xrightarrow{f} \mathcal{G} \xleftarrow{g} B) \otimes (A' \xrightarrow{f'} \mathcal{G'} \xleftarrow{g'} B') \coloneq A + A' \xrightarrow{f+f'} \mathcal{G} + \mathcal{G'} \xleftarrow{g+g'} B + B'  ,
\]
\else
\begin{multline*}
    (A \xrightarrow{f} \mathcal{G} \xleftarrow{g} B) \otimes (A' \xrightarrow{f'} \mathcal{G'} \xleftarrow{g'} B') \coloneq\\ A + A' \xrightarrow{f+f'} \mathcal{G} + \mathcal{G'} \xleftarrow{g+g'} B + B'  , 
\end{multline*}
\fi
and its unit by the initial object of $\mathbb{C}$.
Symmetry is inherited from $\mathbb{C}$ as the coproduct~\cite{MonoidalCoproduct}.
\end{definition}

Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the full sub-category of $\catname{Csp(Hyp(\Sigma))}$ with discrete \textit{ordered} hypergraphs as objects.
\end{definition}

In particular, this means that any morphism in $\HypI{\Sigma}$ is of the form $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$,
where $n,\;m$ are discrete hypergraphs and $\mathcal{G}$ is an arbitrary hypergraph.  The morphisms $f$ and $g$ into $\mathcal{G}$ will specify which vertices are to be considered as inputs and outputs,  respectively.
The ordering of the interface hypergraphs is the way to differentiate between the interpretations of morphisms $f \otimes g$ and $g \otimes f$.

@@

In order to make the interpretation complete with respect to $\catname{S}(\Sigma)$,  it is necessary to restrict our attention to the following subset of hypergraphs with interfaces. 
\begin{definition}[Category of MDA Hypergraphs with Interfaces~\cite{bonchi_string_2022-2}]
\label{def:monogamy_hyp}
We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph;
    \item $f$ and $g$ are monomorphisms;
    \item the in-degree and out-degree of every vertex is at most $1$;
    \item vertices with in-degree $0$ are precisely the image of $f$; and
    \item vertices with out-degree $0$ are precisely the image of $g$.
\end{enumerate}
We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA cospans.
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  We now recall from
\cite{bonchi_string_2022-2} the following theorem,  which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\catname{S}(\Sigma)$. 
\begin{theorem}[Corollary 26 of~\cite{bonchi_string_2022-2}]\label{thm:prop-equiv}
    $\catname{S}(\Sigma) \cong \MdaCospans$
\end{theorem}

\subsection{DPOI-Rewriting for Hypergraphs with Interfaces}
The correspondence of the previous section means that the symmetric monoidal equations of $\catname{S}(\Sigma, \mathcal{E})$ are absorbed in the hypergraph representation.  The equations $\mathcal{E}$ are then to be considered as rewrite rules on string diagrams,  which are then modelled as DPO-rewrites on hypergraphs with interfaces. 
DPO rewriting is a standard technique for formalising graph rewriting categorically. 
In this section we recall standard results of double-pushout (DPO) rewriting of hypergraphs,  as well as the specific notion of DPO rewriting which is sound and complete for arbitrary SMTs,  which is called \textit{convex} rewriting.

The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalised as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square is depicted in Figure~\ref{fig:dpo_dpoi}~(left).

% \[
% \begin{tikzcd}
% \mathcal L \arrow[d, "m"'] & \mathcal K \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
% \mathcal G^{\urcorner}     & \mathcal C \arrow[l] \arrow[r] & ^{\ulcorner} \mathcal H
% \end{tikzcd}
% \]
% CHRIS: WHY IS J DISPLAYING AND NOT R? 
% \[
%     \scalebox{0.8}{
%         \tikzfig{combinatorial_semantics/DPO_square}
%     }
% \]

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI) as shown in Figure~\ref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  
By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  
We now make this identification throughout without further mention.  
Thus we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the rightmost commutative diagram in Figure~\ref{fig:dpo_dpoi} such that the two marked squares are pushouts.
% \[
% \begin{tikzcd}
% \mathcal L \arrow[d, "m"'] & i+j \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
% \mathcal{G}^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner}\mathcal H \\
%                   & n+m \arrow[lu] \arrow[u] \arrow[ru]          &              
% \end{tikzcd}
% \]
\begin{figure}[t!]
    \begin{subfigure}[T]{0.4\linewidth}
    \[
                % \tikzfig{../figures/combinatorial_semantics/DPO_square}
                \begin{tikzcd}
                    {\mathcal{L}} & {\mathcal{K}} & {\mathcal{R}} \\
                    {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}}
                    \arrow["f"', from=1-1, to=2-1]
                    \arrow[from=1-2, to=1-1]
                    \arrow[from=1-2, to=1-3]
                    \arrow[from=1-2, to=2-2]
                    \arrow[from=1-3, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
                    \arrow[from=2-2, to=2-1]
                    \arrow[from=2-2, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
                \end{tikzcd}
    \]
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.4\linewidth}
        \[
    % \scalebox{0.8}{
        % \tikzfig{../figures/combinatorial_semantics/DPOI_square_Hyp}
    % }
    \begin{tikzcd}
        {\mathcal{L}} & {i+j} & {\mathcal{R}} \\
        {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}} \\
        & {n+m}
        \arrow["f"', from=1-1, to=2-1]
        \arrow[from=1-2, to=1-1]
        \arrow[from=1-2, to=1-3]
        \arrow[from=1-2, to=2-2]
        \arrow[from=1-3, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
        \arrow[from=2-2, to=2-1]
        \arrow[from=2-2, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
        \arrow[from=3-2, to=2-1]
        \arrow[from=3-2, to=2-2]
        \arrow[from=3-2, to=2-3]
    \end{tikzcd}
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \captionsetup{belowskip=-3ex}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

\end{definition}
We need to restrict the notions of pushout complement and of match in order to preserve monogamy and directed acyclicity.

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For MDA cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement $i + j \to \mathcal{L}^\bot \to \mathcal G$ as depicted in the square below
% \begin{figure}[h!]
%     \centering
%     \includegraphics[width=0.4\linewidth]{figures/combinatorial_semantics/boundary_complement.png}
% \end{figure}
% https://q.uiver.app/#q=WzAsNSxbMCwwLCJMIl0sWzIsMCwiaStqIl0sWzAsMiwiR157XFx1cmNvcm5lcn0iXSxbMiw0LCJuK20iXSxbMiwyLCJMXntcXGJvdH0iXSxbMyw0LCJbZF8xLGRfMl0iLDJdLFszLDIsIltiXzEsYl8yXSJdLFswLDIsImYiLDJdLFsxLDAsImEgPSBbYV8xLGFfMl0iLDJdLFsxLDQsImMgPSBbY18xLGNfMl0iXSxbNCwyLCJnIl1d
% \[\begin{tikzcd}
% 	\mathcal L && {i+j} \\
% 	\\
% 	{\mathcal G^{\urcorner}} && {\mathcal{L}^{\bot}} \\
% 	\\
% 	&& {n+m}
% 	\arrow["{[d_1,d_2]}"', from=5-3, to=3-3]
% 	\arrow["{[b_1,b_2]}", from=5-3, to=3-1]
% 	\arrow["f"', from=1-1, to=3-1]
% 	\arrow["{[a_1,a_2]}"', from=1-3, to=1-1]
% 	\arrow["{[c_1,c_2]}", from=1-3, to=3-3]
% 	\arrow[from=3-3, to=3-1]
% \end{tikzcd}\]
\[
    \scalebox{0.8}{
        \tikzfig{../figures/combinatorial_semantics/DPOI_boundary_complement}
    }
\]
is called a boundary complement if $[c_1, c_2]$ is mono and there exist $d_1 : n \to \mathcal{L}^{\bot}$ and $d_2 : m \to \mathcal{L}^{\bot}$ making the above triangle commute and such that
\[
    n + j \xrightarrow{[d_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[d_2,c_1]} m + i
\]
is an MDA cospan. 
\end{definition}

Intuitively,  requiring boundary complements ensures that inputs are glued exclusively to outputs and vice-versa in the two pushout squares.
Yet, this restriction is not enough to make DPOI rewriting complete for $\catname{S}(\Sigma,\mathcal{E})$: there might be cospans with carriers $\mathcal{G}$ and $\mathcal{H}$ such that one can be rewritten into another but their corresponding morphisms are not equal modulo $\mathcal{E}$~\cite{bonchi_string_2022-2}.
To make the DPOI rewriting complete, we need to add a restriction on the image of the match.

@@

Because extended cospans have a more general notion of interface,  including \textit{internal} vertices,  DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics} needs some adjustments, defined in this section.

>>>>

First,  we fix some notation.  Let $(-)^*$ be the free monoid monad over \textsf{Set}.  
We extend this to the case of relations $R \subseteq V \times V$,  so that we denote by $R^{*} \subseteq V^* \times V^*$ the element-wise extension of $R$ to a relation on ordered sequences.
We will sometimes use functional notation for relations,  so that given $R \subseteq V \times W$ and $v \in V$,  $R(v) \subseteq W$. 
We elide associativity and unit isomorphisms associated with coproducts $(+)$,  and denote by $\iota_j: X_{j} \rightarrow X_{1} + \ldots X_{j} + \ldots X_{n}$ the $j^{\text{th}}$ injection,  and $[f,g]: A_1 + A_2 \to B$ the co-pairing of two morphisms $f: A_1 \to B$ and $g:A_2 \to B$.
We denote by $A +_{f,g} B$ the pushout of the span $A \xleftarrow{f} C \xrightarrow{g} B$.
We will also refer to $A,B$ as \textit{feet} of the cospan, and to $C$ as \textit{carrier}.
Given an ordered sequence of labels from some alphabet $\Sigma$, $[l_1, \ldots, l_n]$ we will denote with $w([l_1, \ldots, l_n])$ the word concatenated from these labels using $\otimes$, \textit{e.g.}, $w([A,B]) = A \otimes B$.

@@

In ~\cite{bonchi_string_2022-1},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for categorical presentations of SMTs including a Frobenius algebra. 
Hypergraphs generalise graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams (\textit{i.e.}, without requiring a Frobenius algebra) and the category of (cospans of) such hypergraphs.
We then evoke the definition of hypergraphs and corresponding cospans as introduced in~\cite{bonchi_string_2022-1,bonchi_string_2022-2} over not uni-typed signatures.

\begin{definition}[Category of hypergraphs]\label{def:hypergraph}
A \emph{hypergraph $\mathcal{G}$ over monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$} is a tuple $(V,E,s,t,l)$,  where $V$ is a finite set of vertices, $E$ is a finite set of edges, $s : E \to V^{*}$ is a source function, $t : E \to V^{*}$ is a target function,  and $l = (l_{V} : V \to \Sigma_{O},l_{E} : E \to \Sigma_{M})$ is a pair of labelling functions that assigns each edge and vertex an operator and a type from monoidal signature $\Sigma$ respectively.  
The labelling function must respect the typing of the operator: for all edges $e$, such that $l(e) = c : A \to B$,  $l_{V}^{*}(s(e)) = A$ and $l_{V}^{*}(t(e)) = B$.  
We call a hypergraph \textit{discrete} if its set of edges is empty.   
A \emph{hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ is given by a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that the following hold. 
\begin{enumerate}
    \item $\phi_V^*(s_{\mathcal{F}}(e)) = s_{\mathcal{G}}(\phi_E(e))$
    \item $\phi_V^*(t_{\mathcal{F}}(e)) = t_{\mathcal{G}}(\phi_E(e))$
    \item $l_{V,\mathcal{F}}(v) = l_{V,\mathcal{G}}(\phi_V(v))$
    \item $l_{E,\mathcal{F}}(e) = l_{E,\mathcal{G}}(\phi_E(e))$
\end{enumerate}
That is, a homomorphism should preserve sources and targets of edges and preserve the labels.
We denote by $\catname{Hyp(\Sigma)}$ the category of hypergraphs over $\Sigma$ and hypergraph homomorphisms.
\end{definition}
$\catname{Hyp(\Sigma)}$ has all finite colimits and, in particular,  the coproduct is given by the disjoint union of hypergraphs.
The initial object is given by the empty hypergraph.

To properly interpret morphisms of a symmetric monoidal category we need to account for the dangling wires at the diagram's top and bottom that represent inputs and outputs of a morphism.
Such dangling wires are represented as interfaces for a given hypergraph giving rise to a category of cospans of hypergraphs as defined next.

% \begin{definition}[Symmetric monoidal category of cospans]
% Let $\mathbb{C}$ be a category with all finite colimits.  
% A cospan from $X$ to $Y$ is a pair of arrows $X \xrightarrow{} A \xleftarrow{} Y$  in $\mathbb{C}$.  This \textit{category of cospans of $\mathbb{C}$},  denoted $\catname{Csp(\mathbb{C})}$,  has as objects those of $\mathbb{C}$ and as arrows isomorphism classes of cospans.

% Composition of cospans $X \xrightarrow{f} A \xleftarrow{g} Y$ and $Y \xrightarrow{h} B \xleftarrow{i} Z$ is given by pushout: $X \xrightarrow{f} A +_{g,h} B \xleftarrow{i} Z$,  with identity given by the cospan of identities.  Its monoidal product is given by the coproduct of $\mathbb{C}$ as follows: 
% \begin{multline*}
%     (A \xrightarrow{f} \mathcal{G} \xleftarrow{g} B) \otimes (A' \xrightarrow{f'} \mathcal{G'} \xleftarrow{g'} B') \coloneq\\ A + A' \xrightarrow{f+f'} \mathcal{G} + \mathcal{G'} \xleftarrow{g+g'} B + B'  , 
% \end{multline*}
% and its unit by the initial object of $\mathbb{C}$.
% Symmetry is inherited from $\mathbb{C}$ as the coproduct~\cite{MonoidalCoproduct}.
% \end{definition}

% Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
A hypergraph with interfaces is a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\catname{Hyp}(\Sigma)$ where $n$, $m$ are discrete ordered hypergraphs that denote interfaces.
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the category with discrete \textit{ordered} hypergraphs as objects and isomorphism classes of hypergraphs with interfaces as morphisms.
\end{definition}
$\HypI{\Sigma}$ is a symmetric monoidal category with the tensor product given by coproduct in $\catname{Hyp}(\Sigma)$:

\begin{align*}
n \xrightarrow{f} &\mathcal{F} \xleftarrow{g} m\\
&\times \qquad \quad \Coloneqq n + n' \xrightarrow{f + f'} \mathcal{F} + \mathcal{G} \xleftarrow{g + g'} m + m'\\
n' \xrightarrow{f'} &\mathcal{G} \xleftarrow{g'} m'
\end{align*}
Unit is given by the initial object of $\catname{Hyp}(\Sigma)$.
Symmetry is inherited from coproduct in $\catname{Hyp}(\Sigma)$~\cite{MonoidalCoproduct}.

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} k$ is given by the pushout
\[
n \xrightarrow{f} \mathcal{F} +_{g,f'} \mathcal{G} \xleftarrow{g'} k
\]

that is computed by identifying vertices in $\mathcal{F} + \mathcal{G}$ that share pre-image in $m$.
As pushout is only defined up to isomorphism, it justifies the morphisms in this category being isomorphism classes of cospans.

To make such cospans complete with respect to representing morphism of $\catname{S}(\Sigma)$, we need to impose some restriction in their form.

@@

\begin{definition}[Category of MDA Hypergraphs with Interfaces~\cite{bonchi_string_2022-2}]
\label{def:monogamy_hyp}
We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph;
    \item $f$ and $g$ are monomorphisms;
    \item the in-degree and out-degree of every vertex is at most $1$;
    \item vertices with in-degree $0$ are precisely the image of $f$; and
    \item vertices with out-degree $0$ are precisely the image of $g$.
\end{enumerate}
We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA cospans.
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  We now recall from
\cite{bonchi_string_2022-2} the following theorem,  which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\catname{S}(\Sigma)$. 
\begin{theorem}[Corollary 26 of~\cite{bonchi_string_2022-2}]\label{thm:prop-equiv}
    $\catname{S}(\Sigma) \cong \MdaCospans$
\end{theorem}

Practical meaning of the above theorem is that two $\Sigma$-terms $f$ and $g$ are equal modulo SMC equations if and only if they are represented as the same hypergraph with interfaces.
The theorem defines a full and faithful functor that is induced by interpretation function of morphism terms that renders the latter as cospans of hypergraphs and is given in Appendix~\ref{sec:appendix:interpretation}.
Term rewriting and hence equality modulo $\mathcal{E}$ is then implemented as DPO rewrites over such hypergraphs with interfaces that we recall next.

\subsection{DPOI-Rewriting for Hypergraphs with Interfaces}
DPO rewriting is a standard technique for formalising graph transformations categorically.
% In this section we recall standard results of double-pushout (DPO) rewriting of hypergraphs,  as well as the specific notion of DPO rewriting which is sound and complete for arbitrary SMTs,  which is called \textit{convex} rewriting.
The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalised as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square is depicted in Figure~\ref{fig:dpo_dpoi}~(left).

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI) as shown in Figure~\ref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  
By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  
We now make this identification throughout without further mention.  
Thus we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the rightmost commutative diagram in Figure~\ref{fig:dpo_dpoi} such that the two marked squares are pushouts.

\begin{figure}[t!]
    \begin{subfigure}[T]{0.4\linewidth}
    \[
                % \tikzfig{../figures/combinatorial_semantics/DPO_square}
                \begin{tikzcd}
                    {\mathcal{L}} & {\mathcal{K}} & {\mathcal{R}} \\
                    {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}}
                    \arrow["f"', from=1-1, to=2-1]
                    \arrow[from=1-2, to=1-1]
                    \arrow[from=1-2, to=1-3]
                    \arrow[from=1-2, to=2-2]
                    \arrow[from=1-3, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
                    \arrow[from=2-2, to=2-1]
                    \arrow[from=2-2, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
                \end{tikzcd}
    \]
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.4\linewidth}
        \[
    % \scalebox{0.8}{
        % \tikzfig{../figures/combinatorial_semantics/DPOI_square_Hyp}
    % }
    \begin{tikzcd}
        {\mathcal{L}} & {i+j} & {\mathcal{R}} \\
        {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}} \\
        & {n+m}
        \arrow["f"', from=1-1, to=2-1]
        \arrow[from=1-2, to=1-1]
        \arrow[from=1-2, to=1-3]
        \arrow[from=1-2, to=2-2]
        \arrow[from=1-3, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
        \arrow[from=2-2, to=2-1]
        \arrow[from=2-2, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
        \arrow[from=3-2, to=2-1]
        \arrow[from=3-2, to=2-2]
        \arrow[from=3-2, to=2-3]
    \end{tikzcd}
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \captionsetup{belowskip=-3ex}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

\end{definition}
To make DPOI rewriting sound and complete with respect to rewriting of $\Sigma$-terms modulo SMC laws and equations $\mathcal{E}$ the notion of both pushout complement and match must be restricted: otherwise there might be DPO rewrites that are not derivable from SMC laws and $\mathcal{E}$-equations.
This gives rise to the following notion of a \textit{boundary complement} and \textit{convex} DPOI rewriting.

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For MDA cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement $i + j \to \mathcal{L}^\bot \to \mathcal G$ as depicted in the square below
\[
    \adjustbox{scale=0.8}{
        \tikzfig{./figures/DPOI_boundary_complement}
    }\]
is called a boundary complement if $[c_1, c_2]$ is mono and there exist $d_1 : n \to \mathcal{L}^{\bot}$ and $d_2 : m \to \mathcal{L}^{\bot}$ making the above triangle commute and such that
\[
    n + j \xrightarrow{[d_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[d_2,c_1]} m + i
\]
is an MDA cospan. 
\end{definition}

Intuitively,  requiring boundary complements ensures that inputs are glued exclusively to outputs and vice-versa in the two pushout squares.
% Yet, this restriction is not enough to make DPOI rewriting complete for $\catname{S}(\Sigma,\mathcal{E})$: there might be cospans with carriers $\mathcal{G}$ and $\mathcal{H}$ such that one can be rewritten into another but their corresponding morphisms are not equal modulo $\mathcal{E}$~\cite{bonchi_string_2022-2}.
% To make the DPOI rewriting complete, we need to add a restriction on the image of the match.

@@

Because extended cospans have a more general notion of interface,  including \textit{internal} vertices,  DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics} needs some adjustments, defined in this section.
The construction in this section closely follows the one introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} with some adjustments to account for $\textcolor{blue}{<}$.
